---
title: "Explicit ichorCNA inputs, plotting transforms, and run selection"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Explicit ichorCNA inputs, plotting transforms, and run selection}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 8, fig.height = 3)
library(seeNA)
```

This example uses the package's small test-only format fixtures. For manuscript
results, supply your real files and record the exact selected run.

## Read without swallowing warnings

```{r import}
root <- system.file("extdata", package = "seeNA")
x <- read_ichor_sample(
  file.path(root, "example-a.cna.seg"),
  file.path(root, "example-a.seg"),
  file.path(root, "example-a.params.txt"),
  genome_build = "hg38"
)
 c(TF = ichor_tf(x), tumor_ploidy = ichor_ploidy(x))
```

The default `bounds = "window"` clips fixed-width terminal bins that overhang
the reference length, provided the rest of the file sits on a regular bin grid.
Each clipped interval is recorded in `x$coordinate_changes`, and a cohort emits
one summary message rather than one per sample. `bounds = "error"` rejects any
overhang; `bounds = "trim"` clips any straddling interval with a warning.
Rather than wrapping imports in `suppressWarnings()`, handle the classed
conditions `seena_window_padding` (message) and `seena_bounds_warning`
(warning). Window acceptance does not verify the genome build: verify the
assembly from the original run configuration.

## Reuse public helpers rather than duplicate conventions

```{r helpers}
head(ichor_genome_layout("hg38"))
ichor_call_state(c("HOMD", "HETD", "NEUT", "GAIN", "AMP", "HLAMP3", NA))
ichor_state_colors()
ichor_neutral_cn(x)
```

`ichor_neutral_cn()` summarizes observed copy numbers in NEUT bins separately for
autosomes, X and Y, pairing `corrected_call` with `corrected_copy_number` or
`event` with `copy_number`. A value is reported only when exactly one distinct
copy number was observed; otherwise the result is NA with a status and the
candidate values. There is no diploid default or borrowing between chromosome
classes.

## Raw versus upstream-adjusted log ratio

```{r shifted-profile}
plot_ichor_profile(x, ploidy_adjust = TRUE)
```

The opt-in shift matches the **formula** in ichorCNA v0.4
`plotGWSolution() -> plotCNlogRByChr()`:

\[
P_{mixture}=fP_{tumor}+(1-f)2,\qquad
 y=\mathrm{logR}+\log_2(P_{mixture}/2).
\]

`f` is fitted TF as a fraction. The same shift is applied to segment medians, and
`ichor_adjusted_logr(x)` / `ichor_adjusted_logr(x, "segments")` expose the vectors
for custom graphics. Exported parameters are rounded, so the result need not be
pixel-identical to upstream's in-memory plot. The source object is not modified;
plots attach `ichor_transform` metadata. Raw output is the default, and requesting
the adjustment without parameters is an error.

This is a display transform, not integer CN, purity deconvolution, or a new call,
and it does not guarantee that every loss lies below zero or every gain above.

## Cohort missingness is visible, not imputed

```{r cohort}
cohort <- read_ichor_cohort(file.path(root, "example-manifest.csv"), "hg38")
m <- ichor_matrix(cohort, value = "call", chromosomes = "1")
m$values[, 1:4]
m$coverage[, 1:4]
```

ichorCNA filters bins, so a complete reference grid can contain many NA cells
even for valid data. With `min_coverage = 1`, only fully observed target bins
receive a value; inspect `coverage` and `mixed` before lowering the threshold.
Categorical mixtures use the base-pair mode with NA ties, never averages that
could cancel gain and loss to zero.

## Selection support without silently choosing a fitted solution

Multiple fitted runs for the same sample can disagree, and the selected
`.cna.seg`, `.seg` and `.params.txt` must come from one run. The following recipe
(not executed here) compares candidate runs against an externally reported TF.
`candidates` is a user-maintained table with one row per run and complete paths;
`reported_tf` and `tolerance` are fractions, with the tolerance reflecting the
documented reporting precision.

```{r select-run, eval=FALSE}
stopifnot(all(c("run", "cna_seg", "seg", "params") %in% names(candidates)))
stopifnot(!anyDuplicated(candidates$run))
stopifnot(all(file.exists(unlist(candidates[c("cna_seg", "seg", "params")]))))
stopifnot(length(reported_tf) == 1, is.finite(reported_tf),
          reported_tf >= 0, reported_tf <= 1,
          length(tolerance) == 1, is.finite(tolerance), tolerance >= 0)
p <- lapply(candidates$params, read_ichor_params)
source_ids <- vapply(p, function(z) z$sample_id, character(1))
stopifnot(length(unique(source_ids)) == 1)
candidates$fitted_tf <- vapply(p, function(z) z$tumor_fraction, numeric(1))
candidates$delta <- abs(candidates$fitted_tf - reported_tf)
candidates$eligible <- is.finite(candidates$delta) & candidates$delta <= tolerance
candidates  # inspect and save this audit table privately
if (sum(candidates$eligible) != 1L) {
  stop("No unique matching run: resolve absent/ambiguous evidence explicitly.")
}
selected <- candidates[candidates$eligible, ]
x <- read_ichor_sample(selected$cna_seg, selected$seg, selected$params,
                       genome_build = build)
# Save selected run ID, evidence table, tolerance, input fingerprints and commit.
ichor_provenance(x)
```

If several runs match, record the external adjudication rather than silently
choosing one. The package does not fit, rank by biological result, or infer the
correct run.

## Before citing manuscript results

Freeze the package commit and input-run manifest, inspect the exported figures,
and record transformation settings, thresholds, cohort exclusions and plot
scales, including which measurement each heatmap shows. Release and DOI issuance
still need author approval.
