---
title: "Explicit ichorCNA inputs, plotting transforms, and run selection"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Explicit ichorCNA inputs, plotting transforms, and run selection}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 8, fig.height = 3)
library(ichorViz)
```

This example uses the package's small **test-only format fixtures**, not patient
data or a biological simulation. For manuscript results, supply your real files
from an access-controlled location and record the exact selected run.

## Read without swallowing warnings

```{r import}
root <- system.file("extdata", package = "ichorViz")
x <- read_ichor_sample(
  file.path(root, "example-a.cna.seg"),
  file.path(root, "example-a.seg"),
  file.path(root, "example-a.params.txt"),
  genome_build = "hg38"
)
 c(TF = ichor_tf(x), tumor_ploidy = ichor_ploidy(x))
```

The default `bounds = "window"` supports terminal fixed-width windows, including
when upstream filtering has removed intermediate bins. At least two observed
interior bins must support one width; all remaining bin starts/ends must fit the
same 1-based grid, and any segment overhang must end at an observed terminal bin
boundary. Each clipped interval is recorded in `x$coordinate_changes` and import
provenance names the policy. A cohort emits one summary, not one per sample.

`bounds = "error"` remains available for strict interval-bound rejection.
`bounds = "trim"` is a broader opt-in for manually inspected straddling intervals.
Do **not** put `suppressWarnings()` around import. Specific conditions are
`ichorviz_window_padding` (message) and `ichorviz_bounds_warning` (warning).
Other warnings are not swallowed by cohort aggregation.

Window acceptance does not establish the build: hg19 and hg38 can have the same
rounded terminal window on some chromosomes. Verify the assembly from the
original run configuration. Large overhangs are suspicious, but small ones are
not proof of correctness.

## Reuse public helpers rather than duplicate conventions

```{r helpers}
head(ichor_genome_layout("hg38"))
ichor_call_state(c("HOMD", "HETD", "NEUT", "GAIN", "AMP", "HLAMP3", NA))
ichor_state_colors()
ichor_neutral_cn(x)
```

`ichor_neutral_cn()` summarizes observed copy numbers in NEUT bins separately for
autosomes, X and Y. It pairs `corrected_call` with `corrected_copy_number`, or
`event` with `copy_number`. The numeric result is present only with one distinct
observed value. Missing and ambiguous evidence return NA, with counts/status and
candidate values. There is no modal tie break, assumed diploid fallback, or
borrowing between chromosome classes. A supported observed value is still not
proof of an unobserved biological baseline or the run's full calling thresholds.

## Raw versus upstream-adjusted log ratio

```{r shifted-profile}
plot_ichor_profile(x, ploidy_adjust = TRUE)
```

The opt-in shift matches the **formula** in ichorCNA v0.4
`plotGWSolution() -> plotCNlogRByChr()`:

\[
P_{mixture}=fP_{tumor}+(1-f)2,\qquad
 y=\mathrm{logR}+\log_2(P_{mixture}/2).
\]

`f` is fitted TF as a fraction. The same shift is applied to segment medians.
`ichor_adjusted_logr(x)` and `ichor_adjusted_logr(x, "segments")` expose the
vectors for custom graphics. Exported parameters are rounded, so this need not
be pixel-identical to upstream's plot made with full-precision in-memory fits.
The source object is not modified; plotting functions attach `ichor_transform`
metadata. Default `ploidy_adjust = FALSE` deliberately preserves raw output and
supports samples without parameter files. No silent fallback when TRUE lacks
parameters.

This is a display transform, **not** integer CN, purity deconvolution, or a new
call. It does not guarantee every loss will fall below zero or every gain above
zero. Sex-chromosome correction and segment-versus-bin noise can further separate
call labels from y position. Raw segment events and corrected bin states remain
different fields; default grey segments avoid merging their legends.

## Cohort missingness is visible, not imputed

```{r cohort}
cohort <- read_ichor_cohort(file.path(root, "example-manifest.csv"), "hg38")
m <- ichor_matrix(cohort, value = "call", chromosomes = "1")
m$values[, 1:4]
m$coverage[, 1:4]
```

ichorCNA filters bins. A complete reference grid can therefore contain many NA
cells even in valid data; the fraction is dataset- and measurement-specific,
not a fixed package failure rate. With `min_coverage = 1`, only fully observed
target bins receive a value. Inspect `coverage` and `mixed`; lower the coverage
threshold only as an explicit analysis choice. Categorical mixtures use
base-pair mode and NA ties, never averages that can cancel gain and loss to zero.

## Selection support without silently choosing a fitted solution

Multiple fitted runs with the same sample ID can disagree. IDs alone do not
verify run identity; the selected `.cna.seg`, `.seg`, and `.params.txt` paths must
refer to the same run. A published or externally adjudicated TF is useful
selection evidence, not grounds to optimize a manuscript result.

The following **recipe is not executed**: `candidates` is a user-maintained table
of complete paths with one row per explicitly identified run, all for one sample.
`reported_tf` and `tolerance` are externally specified fractions. Tolerance must
reflect documented reporting precision, not be widened until something matches.

```{r select-run, eval=FALSE}
stopifnot(all(c("run", "cna_seg", "seg", "params") %in% names(candidates)))
stopifnot(!anyDuplicated(candidates$run))
stopifnot(all(file.exists(unlist(candidates[c("cna_seg", "seg", "params")]))))
stopifnot(length(reported_tf) == 1, is.finite(reported_tf),
          reported_tf >= 0, reported_tf <= 1,
          length(tolerance) == 1, is.finite(tolerance), tolerance >= 0)
p <- lapply(candidates$params, read_ichor_params)
source_ids <- vapply(p, function(z) z$sample_id, character(1))
stopifnot(length(unique(source_ids)) == 1)
candidates$fitted_tf <- vapply(p, function(z) z$tumor_fraction, numeric(1))
candidates$delta <- abs(candidates$fitted_tf - reported_tf)
candidates$eligible <- is.finite(candidates$delta) & candidates$delta <= tolerance
candidates  # inspect and save this audit table privately
if (sum(candidates$eligible) != 1L) {
  stop("No unique matching run: resolve absent/ambiguous evidence explicitly.")
}
selected <- candidates[candidates$eligible, ]
x <- read_ichor_sample(selected$cna_seg, selected$seg, selected$params,
                       genome_build = build)
# Save selected run ID, evidence table, tolerance, input fingerprints and commit.
ichor_provenance(x)
```

If several runs match the reported TF, selecting the first by directory order is
an additional policy—not proof they are equivalent. Record any external
adjudication rather than silently choosing one. Full files are validated when
the selected sample is read. The package does not fit, rank by biological result,
or infer the correct run.

## Before citing manuscript results

Freeze the package commit and input-run manifest, inspect actual exported
figures, and record transformation settings, thresholds, cohort exclusions and
plot scales. Record whether a heatmap shows raw/adjusted logR, CN, categorical
calls, or a separately derived quantity; CN minus fitted ploidy is not the same
as CN minus an observed NEUT reference. No display dead band establishes a new
neutrality call. Release/archive and DOI issuance still need author approval;
a development version or private repository must not be described as an archived
public software release.
